Large-scale feature selection with Gaussian mixture models for the classification of high dimensional remote sensing images

### 2017-01-12 response.tex

    The paper has been modified:

    \emph{For the sake of clarity, only the three aforementioned configurations are discussed here \rev{and the results for all other configurations is available in the supplementary material}.}
  \end{resbox}
\end{revbox}

    \begin{table}[!t]
        \centering
        \caption{Average classification accuracy 20 trials with all configurations of criterion and method (standard deviation in parenthesis).\label{tab:aisa-otbsimu-all}}
        \begin{tabular}{lccc}\toprule
             & \multicolumn{3}{c}{\bfseries Cohen's kappa} \\ \cmidrule{2-4}
            \# samples by class & 250 & 500 & 1000 \\ \midrule

            GMM SFS JM &      {\bfseries 0.685 (0.030)} & {\bfseries 0.689 (0.030)} & {\bfseries 0.693 (0.032)} \\
            GMM SFS divKL &   {\bfseries 0.680 (0.027)} & {\bfseries 0.699 (0.028)} & {\bfseries 0.708 (0.031)} \\
            GMM SFS kappa &   {\bfseries 0.682 (0.025)} & {\bfseries 0.698 (0.027)} & {\bfseries 0.710 (0.029)} \\
            GMM SFS OA &      {\bfseries 0.682 (0.025)} & {\bfseries 0.698 (0.027)} & {\bfseries 0.710 (0.028)} \\
            GMM SFS F1mean &  {\bfseries 0.682 (0.028)} & {\bfseries 0.699 (0.027)} & {\bfseries 0.710 (0.028)} \\
            GMM SFFS JM &     {\bfseries 0.685 (0.030)} & {\bfseries 0.689 (0.030)} & {\bfseries 0.693 (0.032)} \\
            GMM SFFS divKL &  0.547 (0.158) & 0.615 (0.113) & 0.636 (0.120) \\
            GMM SFFS kappa &  0.519 (0.126) & 0.537 (0.102) & 0.545 (0.141) \\
            GMM SFFS OA &     0.531 (0.126) & 0.537 (0.093) & 0.560 (0.127) \\
            GMM SFFS F1mean & 0.447 (0.158) & 0.579 (0.097) & 0.529 (0.129) \\
            \bottomrule
        \end{tabular}
    \end{table}

    \begin{table}[!t]
        \centering
        \caption{Average classification accuracy for 1,000 samples by class averaged over 5 trials  with all configurations of criterion and method (standard deviation in parenthesis).\label{tab:potsdam-otbsimu-all}}
        \begin{tabular}{lccc}\toprule
            & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries 3\_10 (test)} \\ \cmidrule{2-4}
            GMM SFS JM &      0.624 (0.028) & & \\
            GMM SFS divKL &   0.552 (0.092) & & \\
            GMM SFS kappa &   {\bfseries 0.694 (0.002)} & & \\
            GMM SFS OA &      {\bfseries 0.696 (0.003)} & & \\
            GMM SFS F1mean &  {\bfseries 0.689 (0.002)} & & \\
            GMM SFFS JM &     0.624 (0.028) & & \\
            GMM SFFS divKL &  0.161 (0.066) & & \\
            GMM SFFS kappa &  0.474 (0.103) & & \\
            GMM SFFS OA &     0.407 (0.114) & & \\
            GMM SFFS F1mean & 0.319 (0.052) & & \\
            \bottomrule
        \end{tabular}
    \end{table}

@@

    The paper has been modified:

    \emph{The    classification   rate    is    presented   using    Cohen's kappa \rev{but scores computed with overall accuracy and mean of f1-score are available in supplementary material}.}
  \end{resbox}
\end{revbox}

    \begin{table}[!t]
        \centering
        \caption{Average classification accuracy 20 trials evaluated with overall accuracy and mean of F1-score (standard deviation in parenthesis).\label{tab:aisa-otbsimu-othereval}}
        \begin{tabular}{lcccccc}\toprule
             & \multicolumn{3}{c}{\bfseries Overall Accuracy} & \multicolumn{3}{c}{\bfseries F1-mean} \\ \cmidrule{2-7}
            \# samples by class & 250 & 500 & 1000  & 250 & 500 & 1000 \\ \midrule

            GMM SFS JM &      {\bfseries 0.735 (0.027)} & {\bfseries 0.738 (0.027)} & {\bfseries 0.742 (0.029)} & {\bfseries 0.591 (0.018)} & {\bfseries 0.596 (0.018)} & {\bfseries 0.600 (0.020)} \\
            GMM SFS divKL &   {\bfseries 0.730 (0.024)} & {\bfseries 0.747 (0.026)} & {\bfseries 0.754 (0.029)} & {\bfseries 0.587 (0.017)} & {\bfseries 0.607 (0.017)} & {\bfseries 0.616 (0.019)} \\
            GMM SFS kappa &   {\bfseries 0.733 (0.023)} & {\bfseries 0.746 (0.025)} & {\bfseries 0.756 (0.027)} & {\bfseries 0.591 (0.014)} & {\bfseries 0.607 (0.015)} & {\bfseries 0.620 (0.018)} \\
            GMM SFS OA &      {\bfseries 0.732 (0.023)} & {\bfseries 0.746 (0.025)} & {\bfseries 0.756 (0.026)} & {\bfseries 0.591 (0.014)} & {\bfseries 0.607 (0.015)} & {\bfseries 0.619 (0.018)} \\
            GMM SFS F1mean &  {\bfseries 0.732 (0.026)} & {\bfseries 0.747 (0.025)} & {\bfseries 0.757 (0.026)} & {\bfseries 0.590 (0.018)} & {\bfseries 0.609 (0.016)} & {\bfseries 0.620 (0.018)} \\
            GMM SFFS JM &     {\bfseries 0.735 (0.027)} & {\bfseries 0.738 (0.027)} & {\bfseries 0.742 (0.029)} & {\bfseries 0.591 (0.018)} & {\bfseries 0.596 (0.018)} & {\bfseries 0.600 (0.020)} \\
            GMM SFFS divKL &  0.611 (0.149) & 0.672 (0.105) & 0.690 (0.110) & 0.441 (0.152) & 0.515 (0.114) & 0.541 (0.116) \\
            GMM SFFS kappa &  0.583 (0.119) & 0.599 (0.099) & 0.605 (0.141) & 0.427 (0.117) & 0.439 (0.085) & 0.442 (0.119) \\
            GMM SFFS OA &     0.596 (0.115) & 0.600 (0.090) & 0.620 (0.127) & 0.437 (0.124) & 0.436 (0.079) & 0.455 (0.109) \\
            GMM SFFS F1mean & 0.515 (0.154) & 0.640 (0.093) & 0.590 (0.127) & 0.352 (0.128) & 0.476 (0.085) & 0.431 (0.114) \\
            GMM ridge &       0.665 (0.038) & 0.672 (0.034) & 0.692 (0.033) & {\bfseries 0.580 (0.017)} & {\bfseries 0.593 (0.016)} & {\bfseries 0.609 (0.018)} \\
            KNN &             0.611 (0.033) & 0.622 (0.030) & 0.632 (0.027) & 0.485 (0.019) & 0.502 (0.018) & 0.515 (0.019) \\
            Random Forest &   0.699 (0.023) & 0.723 (0.020) & {\bfseries 0.741 (0.021)} & 0.554 (0.018) & 0.578 (0.017) & {\bfseries 0.598 (0.019)} \\
            \bottomrule
        \end{tabular}
    \end{table}

@@

    \begin{table}[!t]
        \centering
        \caption{Average classification accuracy for 1,000 samples by class averaged over 5 trials  evaluated with overall accuracy and mean of F1-score (standard deviation in parenthesis).\label{tab:potsdam-otbsimu-othereval}}
        \begin{tabular}{lcccccc}\toprule
             & \multicolumn{2}{c}{\bfseries 5\_11 (train)} & \multicolumn{2}{c}{\bfseries 5\_12 (test)} & \multicolumn{2}{c}{\bfseries 3\_10 (test)} \\ \cmidrule{2-7}
             & {\bfseries OA} & {\bfseries F1-mean} & {\bfseries OA} & {\bfseries F1-mean} & {\bfseries OA} & {\bfseries F1-mean} \\ \cmidrule{2-7}

            GMM SFS JM &      0.725 (0.025) & 0.528 (0.022) &  &&&\\
            GMM SFS divKL &   0.677 (0.065) & 0.503 (0.044) &  &&&\\
            GMM SFS kappa &   0.771 (0.002) & 0.642 (0.005) &  &&&\\
            GMM SFS OA &      0.773 (0.002) & 0.642 (0.006) &  &&&\\
            GMM SFS F1mean &  0.767 (0.002) & 0.646 (0.004) &  &&&\\
            GMM SFFS JM &     0.725 (0.025) & 0.528 (0.022) &  &&&\\
            GMM SFFS divKL &  0.408 (0.037) & 0.237 (0.051) &  &&&\\
            GMM SFFS kappa &  0.601 (0.087) & 0.431 (0.065) &  &&&\\
            GMM SFFS OA &     0.544 (0.097) & 0.392 (0.063) &  &&&\\
            GMM SFFS F1mean & 0.473 (0.043) & 0.364 (0.027) &  &&&\\
            \bottomrule
        \end{tabular}
    \end{table}

@@

\section{Reviewer 3}

\begin{revbox}
  Although the method seems promising, there are some significant issues the authors should address to. The key novelty and contribution of this technique is unfortunately unclear. The authors dedicate a lot of space in explaining already known mathematical formulations of the Gaussian Mixture Models and the feature selection techniques, in comparison with their proposed scheme.
  \begin{resbox}
    We totally agree that some mathematical formulations are conventional. We believe that it is necessary to provide to readers sufficient information to explain clearly the contribution of the proposed paper. The contributions are based on specific derivation of the GMM and the FE formulation. Without these primary recall, we are afraid that readers can be lost. Concerning the contribution of the paper, they are stated in the antepenultimate paragraph of the introduction. They concern the expression of efficient updates rules using block matrix processing. Furthermore, an open source C++ library is made available for the community. The results of our experiment shows that the proposed algorithm has the advantage to be very efficient in term of processing time with a fairly reasonable cost in term of classification accuracy. It has to be noticed that compared to similar methods using Gaussian Mixture model, no additional hypothesis are made on the covariance matrix which is tractable only because of the efficient implementation.

    To make the contribution more clear, the antepenultimate paragraph has been modified to

\rev{The contribution of the paper is a extension of the forward feature selection method proposed in~[27]. An smart implementation of the feature selection updates rules are presented in order to perform efficiently on large amount of data. They are based on linear algebra on block matrix applied to the covariance matrix of the conditional class density function.  Furthermore, a floating version of the algorithm is proposed and evaluated. Several \emph{goodness of fit} criteria are proposed to handle unbalanced training set, extending the conventional \emph{overall accuracy} measure.  Finally, the developed algorithm is made available to the scientific community as a remote module of the Orfeo Toolbox~[13].}
  \end{resbox}
\end{revbox}

### 2017-01-12 paper.tex
improve english

% \tableofcontents
% \clearpage
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
  A  large  scale  feature  selection wrapper  is  discussed  for  the
  classification  of high  dimensional  remote  sensing. An  efficient
  implementation is proposed based on intrinsic properties of Gaussian
  mixtures models and  block matrix.  The criterion  function is split
  into two parts  : one that is  updated to test each  feature and one
  that needs to be updated only once per feature selection. This split
  saved  a  lot  of  computation  for each  test.   The  algorithm  is
  implemented in  \texttt{C++} and integrated into  the Orfeo Toolbox.
  It has been compared to  other classification algorithms on two high
  dimension remote sensing images. The  results show that the approach
  provide good classification accuracies with low computation time.
\end{abstract}

@@

\IEEEPARstart{W}{ith} the increasing number of remote sensing missions, the quantity of available Earth observation data for a given landscape becomes larger and larger. Satellite missions produce a huge amount of data on a regularly (daily) basis. From 2018, the EnMAP (Environmental Mapping and Analysis Program) satellites managed by the German space agency will produce images with 244 spectral bands, a spatial resolution of 30x30m per pixel and with a frequency of revisit of 4 days\cite{Müller09enmap}. The Hyperspectral Infrared Imager (HyspIRI) of NASA will also deliver images with 212 spectral bands. Additionally to hyperspectral data, the amount of available hypertemporal data increases a lot. For instance, the European satellites Sentinel-2 were launched recently and 2 Terabytes of data will be released every day~\cite{drusch2012sentinel}. The Landsat open archive (\url{http://landsat.usgs.gov/products_data_at_no_charge.php}) has released thousands of images. Such high volume of Earth observation data provides an accurate information of land state and functioning, and helps to improve the understanding of the planet~\cite{rs1010001}. However, processing such data is more and more challenging because of statistical and computational issues.

In the spectral or temporal domain, a pixel is represented by a vector
for   which  each   component  corresponds   to  a   spectral/temporal
measurement.   The size  of  the  vector is  therefore  the number  of
spectral  or temporal  measurements.  For  hyperspectral images,  this
number is  typically about several  hundreds while for  the Sentinel-2
multitemporal images, the number of spectro-temporal measurement for a
given  year  is approximately  one  thousand.   When working  in  high
dimensional  space,  statistical  methods  made for  low  or  moderate
dimensional  space do  not  adapt  well.  For  instance,  the rate  of
convergence of the statistical estimation decreases when the dimension
grows while conjointly the number of parameters to estimate increases,
making    the    estimation    of   the    model    parameters    very
difficult~\cite{donoho}.  Consequently,  with a limited  training set,
beyond a certain limit, the classification accuracy actually decreases
as the number of features increases~\cite{hughes}.  For the purpose of
classification,  these  problems are  related  to  the \emph{curse  of
  dimensionality}~\cite{donoho}.  This is  a  major  drawback in  many
remote sensing  application since it  is difficult to collect  a large
and an accurate ground-truth.  An intensive work has been performed in
the remote  sensing community to  build accurate classifiers  for high
dimensional  images.   Bayesian models~\cite{book:landgrebe},  feature
extraction              and             feature              reduction
techniques~\cite{book:landgrebe,DR:guided:tour},                random
forest~\cite{1396322},   neural  networks~\cite{5411821}   and  kernel
methods~\cite{kernel:methods:rs}  have   been  investigated   for  the
classification of such images.

The volume of the data is  increasing dramatically with respect to the
number of measurement  per pixel. The data volume  of an hyperspectral
image  is  typically several  hundreds  of  Gigabytes per  acquisition
($\approx$  300km$^2$). Multitemporal  data are  now available  freely
from  internet  stream  (see  for instance  the  Copernicus  data  hub
\url{https://cophub.copernicus.eu/}).  This very  large volume of data
requires   specific   computing  infrastructure.    High   performance
computing   is   actually   investigated   by   the   remote   sensing
community~\cite{christophe2011remote,plaza2011high}.  Main  issues are
related to the use of parallel approach (multi-core, GPU, clusters) to
improve the  processing time, and  to the use of  streaming techniques
when data  does not fit in  memory.  One popular open  source software
solution  is  the Orfeo  Toolbox,  developed  by the  French  Space
Agency~\cite{christophe2008orfeo}.   Streaming and  parallel computing
are  conveniently proposed  to the  users/developpers through  several
modules ``\emph{ready to use}''.

A method  to reduce  both statistical and  computational issues  is to
perform a reduction of the dimension. In fact, with the \emph{curse of
  dimensionality}     comes     the      \emph{blessing     of     the
  dimensionality}~\cite{bouveyron2014model}:  high   dimensional  data
space   exhibits  interesting   properties   for  the   classification
purpose.  In  particular,  it  is   possible  to  get  a  parsimonious
representation  of  the  data  while  maintaining  or  increasing  the
classification  accuracy~\cite{jimenez1998supervised}.  For  instance,
in land-cover  classification, given  a set  of spatial,  temporal and
spectral features, it is possible to  extract those which are the most
discriminant      for      the     purpose      of      classification
\cite{fassnacht2014comparison}.   In  hyperspectral   data,  from  the
hundreds of available spectral channels,  it is possible to reduce the
number of channels  to make the processing more efficient  in terms of
statistical complexity  and computational load. In  short, by reducing
the dimension better, classification results is expected with a reduced
computational load.

>>>>

% \tableofcontents
% \clearpage
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
  A  large  scale  feature  selection wrapper  is  discussed  for  the
  classification  of high  dimensional  remote  sensing. An  efficient
  implementation is proposed based on intrinsic properties of Gaussian
  mixtures models and  block matrix.  The criterion  function is split
  into two parts  : one that is  updated to test each  feature and one
  that needs to be updated only once per feature selection. This split
  saved  a  lot  of  computation  for each  test.   The  algorithm  is
  implemented in  \texttt{C++} and integrated into  the Orfeo Toolbox.
  It has been compared to  other classification algorithms on two high
  dimension remote sensing images. The  results show that the approach
  provides good classification accuracies with low computation time.
\end{abstract}

@@

\IEEEPARstart{W}{ith} the increasing number of remote sensing missions, the quantity of available Earth observation data for a given landscape becomes larger and larger. Satellite missions produce a huge amount of data on a regularly (daily) basis. From 2018, the EnMAP (Environmental Mapping and Analysis Program) satellites managed by the German space agency will produce images with 244 spectral bands, a spatial resolution of 30x30m per pixel and with a frequency of revisit of 4 days\cite{Müller09enmap}. The Hyperspectral Infrared Imager (HyspIRI) of NASA will also deliver images with 212 spectral bands. Additionally to hyperspectral data, the amount of available hypertemporal data increases a lot. For instance, the European satellites Sentinel-2 were launched recently and 2 Terabytes of data will be released every day~\cite{drusch2012sentinel}. The Landsat open archive (\url{http://landsat.usgs.gov/products_data_at_no_charge.php}) has also released thousands of images. Such high volume of Earth observation data provides accurate information of land state and functions, and helps to improve the understanding of the planet~\cite{rs1010001}. However, processing such data is more and more challenging because of statistical and computational issues.

In the spectral or temporal domain, a pixel is represented by a vector
for   which  each   component  corresponds   to  a   spectral/temporal
measurement.   The size  of  the  vector is  therefore  the number  of
spectral  or temporal  measurements.  For  hyperspectral images,  this
number is  typically about several  hundreds while for  the Sentinel-2
multitemporal images, the number of spectro-temporal measurement for a
given  year  is approximately  one  thousand.   When working  in  high
dimensional  spaces,  statistical  methods  made for  low  or  moderate
dimensional  spaces do  not  adapt  well.  For  instance,  the rate  of
convergence of the statistical estimation decreases when the dimension
grows while jointly the number of parameters to estimate increases,
making    the    estimation    of   the    model    parameters    very
difficult~\cite{donoho}.  Consequently,  with a limited  training set,
beyond a certain limit, the classification accuracy actually decreases
as the number of features increases~\cite{hughes}.  For the purpose of
classification,  these  problems are  related  to  the \emph{curse  of
  dimensionality}~\cite{donoho}.  This is  a  major  drawback in  many
remote sensing  applications since it  is difficult to collect  a large
and accurate ground-truth.  An intensive work has been performed in
the remote  sensing community to  build accurate classifiers  for high
dimensional  images.   Bayesian models~\cite{book:landgrebe},  feature
extraction              and             feature              reduction
techniques~\cite{book:landgrebe,DR:guided:tour},                random
forest~\cite{1396322},   neural  networks~\cite{5411821}   and  kernel
methods~\cite{kernel:methods:rs}  have   been  investigated   for  the
classification of such images.

The volume of the data is  increasing dramatically with respect to the
number of measurement  per pixel. The data volume  of an hyperspectral
image  is  typically several  hundreds  of  Gigabytes per  acquisition
($\approx$  300km$^2$). Multitemporal  data are  now available  freely
from  internet  stream  (see  for instance  the  Copernicus  data  hub
\url{https://cophub.copernicus.eu/}).  This very  large volume of data
requires   specific   computing  infrastructure.    High   performance
computing   is   actually   investigated   by   the   remote   sensing
community~\cite{christophe2011remote,plaza2011high}.  Main  issues are
related to the use of parallel approaches (multi-core, GPU, clusters) to
improve the  processing time, and  to the use of  streaming techniques
when data  does not fit in  memory.  One popular open  source software
solution  is  the Orfeo  Toolbox,  developed  by the  French  Space
Agency~\cite{christophe2008orfeo}.   Streaming and  parallel computing
are  conveniently proposed  to  users/developpers through  several
modules ``\emph{ready to use}''.

A method  to reduce  both statistical and  computational issues  is to
perform a reduction of the dimension. In fact, with the \emph{curse of
  dimensionality}     comes     the      \emph{blessing     of     the
  dimensionality}~\cite{bouveyron2014model}:  high   dimensional  data
spaces   exhibit  interesting   properties   for    classification
purpose.  In  particular,  it  is   possible  to  get  a  parsimonious
representation  of  the  data  while  maintaining  or  increasing  the
classification  accuracy~\cite{jimenez1998supervised}.  For  instance,
in land-cover  classification, given  a set  of spatial,  temporal and
spectral features, it is possible to  extract those which are the most
discriminant      for      the     purpose      of      classification
\cite{fassnacht2014comparison}.   In  hyperspectral   data,  from  the
hundreds of available spectral channels,  it is possible to reduce the
number of channels  to make the processing more efficient  in terms of
statistical complexity  and computational load. In  short, by reducing
the dimension, better classification results are expected with a reduced
computational load.
